\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}
\usepackage{color}
\usepackage{hyperref}
\usepackage{microtype}
\renewcommand\UrlFont{\color{blue}\rmfamily}

\title{Explainable Quantum-Inspired Multi-label Classification of Non-Functional Requirements}
\titlerunning{Explainable Quantum-Inspired Multi-label NFR Classification}

\author{Anonymous Author(s)}
\authorrunning{Anonymous}

\institute{Affiliation withheld for double-blind review}

\begin{document}
\maketitle

\begin{abstract}
Non-functional requirements (NFRs) frequently overlap across quality attributes
such as security, performance, usability, and availability, making NFR
classification a natural multi-label problem. This paper presents an
intrinsically explainable quantum-inspired method for multi-label NFR
classification. Requirements are represented as normalized semantic states,
NFR labels as projection directions, and label scores as squared projection
amplitudes; a label interference matrix models NFR co-occurrence, and
token-level amplitude contributions explain predictions without a post-hoc
explainer. The paper also gives a theoretical characterization of the deployed
decision rule, identifying when the quantum-inspired geometry reduces to
cosine-style ranking and when interference or row-wise normalization creates
genuine cross-label coupling. We evaluate quantum-inspired variants on the
NICE multi-label dataset against TF-IDF Logistic Regression, TF-IDF Linear SVM,
Sentence-BERT, fine-tuned DistilBERT, and a label-frequency baseline. Under
5-fold cross-validation, Sentence-BERT achieves the highest Macro-F1; the
hybrid quantum--SVM model is competitive with TF-IDF baselines and obtains the
highest Macro-F1 among non-embedding models, but its advantage is not
statistically significant at this dataset size. The main contribution is
therefore not a state-of-the-art accuracy claim, but an interpretable
multi-label scoring formulation, a precise account of when its geometry helps
or collapses to classical behavior, and a deletion-based faithfulness
evaluation of its intrinsic token explanations.
\keywords{Requirements Engineering \and Non-Functional Requirements \and Multi-label Classification \and Quantum-inspired Model \and Explainable AI \and Software Engineering}
\end{abstract}

\section{Introduction}
Requirements Engineering (RE) is a critical activity in software development
because requirement quality strongly affects downstream architecture,
implementation, testing, and maintenance decisions. Among requirement types,
non-functional requirements (NFRs) are particularly difficult to identify and
classify automatically. NFRs describe quality attributes such as security,
performance, usability, availability, scalability, and maintainability. Unlike
many functional requirements, NFRs are often ambiguous, context-dependent, and
overlapping.

For example, a requirement stating that payment information must be encrypted
and accessible only to authorized administrators clearly relates to security. If
the same requirement also constrains availability, auditability, or response
time, it may express multiple NFR categories at once. Therefore, NFR
classification is better treated as a multi-label problem rather than a strict
single-label problem \cite{tsoumakas2007multilabel,zhang2014review}.

This paper investigates the research topic: \emph{Explainable Quantum-Inspired
Multi-label Classification of Non-Functional Requirements}. The key idea is to
represent each requirement as a semantic state and estimate its degree of
association with multiple NFR labels using projection amplitudes. The approach
is quantum-inspired rather than quantum-hardware-based: it runs on ordinary
computers and borrows mathematical intuitions such as state representation,
projection amplitude, and interference from geometric information retrieval and
quantum language modeling \cite{vanrijsbergen2004geometry,sordoni2013quantum,li2019cnm}.

The paper's positive claim is that this formulation provides an intrinsic
explanation mechanism for multi-label NFR classification and a testable
theory of when quantum-inspired geometry changes the decision rule. The basic
projection component is closely related to centroid-based text classification;
the interference and row-wise normalization analyses make explicit when the
model becomes more than an independent centroid classifier and when it does
not. Empirically, the hybrid model offers a small, suggestive accuracy gain
over comparable TF-IDF baselines, while the primary value of the method lies in
its transparent scoring structure and diagnostic ablation.

The contributions of this paper are:
\begin{itemize}
    \item an intrinsically explainable quantum-inspired formulation for multi-label NFR classification;
    \item a theoretical characterization of interference and row-wise normalization as the two sources of cross-label coupling;
    \item token-level semantic amplitude explanations evaluated with a deletion-based ERASER faithfulness test;
    \item empirical comparison against TF-IDF Logistic Regression, TF-IDF Linear SVM, Sentence-BERT, fine-tuned DistilBERT, and a label-frequency baseline on the NICE dataset;
    \item an ablation study showing where projection type, label interference, and hybrid fusion weight help or fail.
\end{itemize}

\section{Related Work}
Automated NFR classification has been studied for many years in Requirements
Engineering. Early work by Cleland-Huang et al. explored automated
classification of NFRs such as security, performance, and usability
\cite{clelandhuang2007automated}. Later studies introduced supervised machine
learning and dependency-based explanations for requirements classification
\cite{kurtanovic2017automatically,dalpiaz2019requirements}. The PROMISE and
PROMISE-expanded datasets have supported repeatable experiments on requirement
classification \cite{mitrevski_promise_exp}.

NICE introduced a multi-label NFR dataset based on the augmented PROMISE dataset
and relabeled NFR sub-classes such as availability, security, and performance
using a multi-label schema \cite{nice_dataset,nice_preprint}. This dataset is
especially relevant to the present work because the research problem is
explicitly multi-label. A direct numeric comparison with the NICE paper is
limited by differences in label filtering, splits, and evaluation protocol, so
we report our own baselines on the same downloaded data.

Traditional text classification commonly relies on bag-of-words or TF-IDF
features combined with classifiers such as Logistic Regression or Support Vector
Machines \cite{sebastiani2002machine}. Transformer-based models have become
strong baselines for requirements classification, including BERT-style transfer
learning approaches such as NoRBERT \cite{devlin2019bert,hey2020norbert}.
Sentence-BERT provides efficient sentence embeddings for downstream classifiers
\cite{reimers2019sentencebert}, and is therefore included as a stronger
embedding baseline in this study.

The quantum-inspired component is related to geometric information retrieval and
quantum-inspired language models, where term dependencies and matching can be
modeled through vector spaces, projections, and complex-valued or
wave-function-inspired representations
\cite{vanrijsbergen2004geometry,sordoni2013quantum,zhang2018quantum,li2019cnm}.
For explainability, this work is positioned near model explanation methods such
as LIME and SHAP \cite{ribeiro2016lime,lundberg2017shap}, and evaluates
faithfulness using a deletion test inspired by rationale evaluation work such
as ERASER \cite{deyoung2020eraser}.

\section{Research Questions}
This study is guided by three research questions:

\textbf{RQ1.} Can a quantum-inspired semantic projection model classify
multi-label NFRs competitively compared with classical and embedding baselines?

\textbf{RQ2.} Does the model provide useful explainability through token-level
semantic amplitude contributions?

\textbf{RQ3.} Is the observed behavior stable under cross-validation rather than
only in a single train/test split?

\section{Proposed Method}
\subsection{Requirement Representation}
Each requirement text is first converted into a TF-IDF vector:
\begin{equation}
    \mathbf{x}_i \in \mathbb{R}^{d},
\end{equation}
where $d$ is the vocabulary size. The vector is normalized to form a semantic
state:
\begin{equation}
    \lvert r_i \rangle = \frac{\mathbf{x}_i}{\left\|\mathbf{x}_i\right\|_2}.
\end{equation}
This normalization allows the requirement to be interpreted as a direction in
semantic space.

\subsection{Label Projection}
For each NFR label $c$, a label basis vector is estimated from the centroid of
positive training examples:
\begin{equation}
    \lvert c \rangle = \frac{1}{N_c}\sum_{i:y_{ic}=1}\lvert r_i \rangle,
\end{equation}
followed by normalization. The raw association between requirement $r_i$ and
label $c$ is computed as a squared projection amplitude:
\begin{equation}
    s_c(r_i) = \left(\langle c \mid r_i \rangle\right)^2.
\end{equation}
We call $s_c(r_i)$ an association score rather than a probability because the
scores over labels are not constrained to sum to one.

\subsection{Contrastive Label Projection}
The positive-only centroid projection is simple and interpretable, but it does
not explicitly model how a label differs from other labels. Therefore, this
work also evaluates a contrastive quantum-inspired variant. For each label $c$,
the contrastive direction is computed as:
\begin{equation}
    \lvert c_{\Delta} \rangle =
    \frac{\mu_c^{+} - \mu_c^{-}}
    {\left\|\mu_c^{+} - \mu_c^{-}\right\|_2},
\end{equation}
where $\mu_c^{+}$ is the centroid of requirements that contain label $c$, and
$\mu_c^{-}$ is the centroid of requirements that do not contain label $c$.
Unlike the positive centroid vector, the contrastive direction is signed: a
requirement can project in the opposite direction of a label. Squaring a
negative projection would incorrectly assign a high score to evidence against
the label. Therefore, the contrastive variant uses a half-wave rectified
projection score:
\begin{equation}
    s^{\Delta}_c(r_i) =
    \left(\max(0,\langle c_{\Delta}\mid r_i\rangle)\right)^2.
\end{equation}
This preserves positive alignment with the label direction while suppressing
opposite-direction evidence.

\subsection{Interference Adjustment}
To reflect label co-occurrence, the model estimates a label interference matrix
from training labels. Let $N_{kl}$ be the number of training samples containing
both labels $k$ and $l$, and let $N_k$ be the number of samples containing label
$k$. The normalized co-occurrence matrix is:
\begin{equation}
    M_{kl} = \frac{N_{kl}}{N_k},
\end{equation}
with the diagonal set to 1. The implementation mixes this matrix with the
identity matrix:
\begin{equation}
    \mathbf{M}' = (1-\lambda)\mathbf{I} + \lambda\mathbf{M},
\end{equation}
where $\lambda$ controls interference strength. For projection scores
$\mathbf{p}_i$, adjusted scores are computed as:
\begin{equation}
    \mathbf{s}_i = \mathrm{normalize}(\mathbf{p}_i \mathbf{M}').
\end{equation}
For the positive-projection model, $\mathrm{normalize}(\cdot)$ denotes row-wise
min--max normalization into $[0,1]$. For the contrastive model, the rectified
projection is passed through a sigmoid calibration and the interference-adjusted
scores are clipped into $[0,1]$. The resulting scores are association scores
for thresholding rather than mutually exclusive probabilities.

\subsection{Hybrid Quantum--SVM Fusion}
To improve stability on small and imbalanced datasets, a hybrid model combines
the contrastive quantum-inspired score with a one-vs-rest Linear SVM score:
\begin{equation}
    \mathbf{h}_i = \alpha \mathbf{q}_i + (1-\alpha)\mathbf{v}_i,
\end{equation}
where $\mathbf{q}_i$ is the contrastive quantum-inspired score vector,
$\mathbf{v}_i$ is the calibrated SVM score vector, and $\alpha$ is the quantum
fusion weight. In the final experiments, $\alpha=0.30$ is used as a moderate
quantum weight; the ablation study (Table~\ref{tab:nice-ablation}) shows that
Macro-F1 is not sensitive to the exact value in this range ($\alpha=0.15$,
$0.30$, and $0.50$ are within one standard deviation of each other), so
$\alpha=0.30$ was kept as a round, non-edge value rather than as a value
selected for maximizing cross-validation Macro-F1.

\subsection{Explainability}
For a predicted label $c$, token-level contribution is estimated from the
element-wise product between the requirement state and the label basis:
\begin{equation}
    e_{ijc} = r_{ij}c_j.
\end{equation}
Terms with the largest absolute contributions are reported as explanations. In
the experiments, these explanations are generated by the rectified contrastive
projection model. We evaluate them using a deletion test: if the highlighted
terms are faithful, deleting them should reduce the model score more than
deleting random nonzero terms.

\subsection{Theoretical Characterization}
The proposed formulation exposes two independent sources of cross-label
coupling in the deployed decision rule: the interference matrix
(Proposition~\ref{prop:interference}) and the row-wise score normalization
that the implementation always applies before thresholding
(Proposition~\ref{prop:reduction}). Only the within-instance \emph{ranking}
of labels reduces to a cosine-similarity ranking; the thresholded binary
decision does not reduce to $L$ independent per-label centroid classifiers,
even when interference is switched off.

\begin{proposition}[Interference as label coupling]
\label{prop:interference}
With $\mathbf{M}'=(1-\lambda)\mathbf{I}+\lambda\mathbf{M}$, the pre-normalization
score for label $l$ has the form
\begin{equation}
    u_l \propto (1-\lambda)p_l + \lambda\sum_k p_k M_{kl}.
\end{equation}
The second term injects score mass from co-occurring labels $k$, where
$M_{kl}=N_{kl}/N_k$. This coupling cannot be represented by a diagonal
per-label centroid classifier; together with the row-wise normalization in
Proposition~\ref{prop:reduction}, it is one of two mechanisms by which the
deployed decision rule departs from an independent per-label classifier.
\end{proposition}

\begin{proof}
Expanding $\mathbf{p}\mathbf{M}'$ gives
$\mathbf{p}((1-\lambda)\mathbf{I}+\lambda\mathbf{M})$, so the $l$-th component
is $(1-\lambda)p_l+\lambda\sum_k p_kM_{kl}$. If $\mathbf{M}'$ were diagonal,
each adjusted label score would depend only on its own projection score $p_l$.
The off-diagonal entries $M_{kl}$ for $k\neq l$ therefore introduce explicit
cross-label coupling through co-occurrence statistics.
\end{proof}

\begin{proposition}[Ranking invariance under row-wise normalization]
\label{prop:reduction}
Let $\mathbf{u}_i = \mathbf{p}_i\mathbf{M}'$ be the pre-normalization score
vector from Proposition~\ref{prop:interference}, and let
$\mathbf{s}_i=\mathrm{normalize}(\mathbf{u}_i)$ (row-wise min--max into
$[0,1]$) be the score vector actually thresholded by the implementation. For
any two labels $k,l$ evaluated on the same instance $i$,
$$s_{i,k}\geq s_{i,l} \iff u_{i,k}\geq u_{i,l};$$
in particular, at $\lambda=0$ (so $u_{i,l}=p_{i,l}=\langle l\mid r_i\rangle^2$),
this ranking equals the ranking by cosine similarity
$\langle l\mid r_i\rangle$. However, for a fixed global threshold
$\tau\in(0,1)$, the decision $s_{i,l}\geq\tau$ is equivalent to
$$u_{i,l} \;\geq\; \min_k u_{i,k} + \tau\Bigl(\max_k u_{i,k}-\min_k u_{i,k}\Bigr) =: \tau_i,$$
where $\tau_i$ depends on instance $i$ through the scores of \emph{every}
label on that instance, not only label $l$. Consequently, $\tau$ does not
correspond to a single instance-independent cutoff on $u_{i,l}$ (or, at
$\lambda=0$, on cosine similarity): two instances with an identical raw
score for label $l$ can receive opposite decisions for that label if their
scores on other labels differ.
\end{proposition}

\begin{proof}
For fixed $i$ with $\max_k u_{i,k} > \min_k u_{i,k}$,
$s_{i,l}=(u_{i,l}-\min_k u_{i,k})/(\max_k u_{i,k}-\min_k u_{i,k})$ is a
strictly increasing affine function of $u_{i,l}$ with the same slope and
intercept for every label $l$ of instance $i$, which proves the
ranking-invariance claim. At $\lambda=0$, $u_{i,l}=\langle l\mid r_i\rangle^2$,
and since $\langle l\mid r_i\rangle=\cos\theta_{r_i,l}\geq 0$ (TF-IDF features
and positive centroids are non-negative), squaring is order-preserving on
$[0,\infty)$, so the ranking by $u_{i,l}$ equals the ranking by
$\cos\theta_{r_i,l}$. Solving $s_{i,l}\geq\tau$ for $u_{i,l}$ using the
definition of $s_{i,l}$ gives the threshold $\tau_i$ above, which is a
function of $\min_k u_{i,k}$ and $\max_k u_{i,k}$ and therefore of the scores
of labels $k\neq l$ on instance $i$; two instances with equal $u_{i,l}$ but
different label-score profiles can therefore yield different values of
$\tau_i$ relative to $u_{i,l}$, hence different decisions. When
$\max_k u_{i,k}=\min_k u_{i,k}$, the implementation sets $s_{i,l}=0$ for all
$l$, so no label is predicted for instance $i$ at any $\tau>0$.
\end{proof}

\begin{remark}
For two labels and $\tau=0.5$, $\mathbf{u}=(0.5,0.0)$ predicts label 1 after
normalization, whereas $\mathbf{u}=(0.5,0.9)$ gives the same raw label-1 score
but does not predict label 1. Row-wise normalization is therefore a second
source of cross-label coupling, independent of $\mathbf{M}'$.
\end{remark}

\section{Experimental Setup}
\subsection{Datasets}
\textbf{NICE multi-label dataset.} The main experiment uses the NICE dataset
from Zenodo \cite{nice_dataset}. The downloaded CSV contains 622 requirements
and binary label columns for NFR classes. After filtering rows with at least one
NFR sub-class label and removing the empty ``Other'' label, the experiment uses
381 requirements and 11 NFR labels. Among these, 125 requirements contain more
than one NFR label, giving a label cardinality of 1.3648.

\textbf{PROMISE-expanded dataset.} A secondary experiment uses
PROMISE-expanded from the public requirements classification repository
\cite{mitrevski_promise_exp}. This dataset is treated as an 11-class
single-label NFR subtype classification task and is used as an auxiliary sanity
check rather than the main multi-label experiment.

\subsection{Compared Models}
The compared baselines are:
\begin{itemize}
    \item \textbf{Label-frequency baseline:} predicts labels according to training label frequencies.
    \item \textbf{TF-IDF Logistic Regression:} one-vs-rest Logistic Regression using unigram and bigram TF-IDF features.
    \item \textbf{TF-IDF Linear SVM:} one-vs-rest Linear SVM using unigram and bigram TF-IDF features.
    \item \textbf{Sentence-BERT + LR:} all-MiniLM-L6-v2 embeddings followed by one-vs-rest Logistic Regression.
    \item \textbf{Fine-tuned DistilBERT:} distilbert-base-uncased fine-tuned directly for multi-label classification with BCE loss and per-label threshold calibration.
\end{itemize}

The quantum-inspired models are:
\begin{itemize}
    \item \textbf{Quantum-inspired Projection:} positive-centroid semantic projection with label interference.
    \item \textbf{Contrastive Quantum Projection:} positive-minus-negative semantic projection with label interference.
    \item \textbf{Hybrid Quantum--SVM Fusion:} score-level fusion of contrastive quantum projection and Linear SVM, using a quantum fusion weight of 0.30.
\end{itemize}

\subsection{Metrics and Threshold Selection}
The main reported metrics are Micro-F1, Macro-F1, Hamming loss, and Label
Ranking Average Precision (LRAP). Macro-F1 is emphasized because the NFR labels
are imbalanced. Two threshold protocols are evaluated. First, a single global
threshold is selected on an internal validation split by maximizing Macro-F1
over thresholds from 0.05 to 0.95. Second, a per-label threshold protocol
selects an independent threshold for each NFR label on the validation split.

\subsection{Hyperparameters}
Table \ref{tab:hyperparameters} summarizes the main hyperparameters used for
reproducibility.

\begin{table}[htbp]
\centering
\caption{Main hyperparameters used in the experiments.}
\label{tab:hyperparameters}
\resizebox{\textwidth}{!}{
\begin{tabular}{ll}
\toprule
Component & Setting \\
\midrule
TF-IDF features & unigrams+bigrams, min\_df=1 \\
Sublinear TF-IDF & enabled for contrastive, hybrid, and SVM-only ablation \\
Logistic Regression & one-vs-rest, class\_weight=balanced, max\_iter=3000 \\
Linear SVM & one-vs-rest LinearSVC, class\_weight=balanced \\
Sentence-BERT & all-MiniLM-L6-v2 + one-vs-rest Logistic Regression \\
Fine-tuned DistilBERT & 2 epochs, batch size 16, max length 128, learning rate $5\times10^{-5}$ \\
Positive projection interference & $\lambda=0.15$ \\
Contrastive projection interference & $\lambda=0.05$ \\
Hybrid projection interference & $\lambda=0.03$ \\
Hybrid quantum weight & $\alpha=0.30$ unless varied in ablation \\
Threshold grid & 0.05 to 0.95 with step 0.05 \\
Random seed & 42 \\
\bottomrule
\end{tabular}
}
\end{table}

\section{Results}
\subsection{Single Train/Validation/Test Split on NICE}
Table \ref{tab:nice-single} reports results from a single
train/validation/test split on the NICE dataset.

\begin{table}[htbp]
\centering
\caption{NICE multi-label results on a single train/validation/test split, comparing label-frequency, TF-IDF, Sentence-BERT, and quantum-inspired models across Micro-F1, Macro-F1, Hamming loss, and LRAP. Sentence-BERT has the highest Macro-F1 and LRAP on this split; the hybrid quantum--SVM model and the contrastive quantum projection share the lowest Hamming loss.}
\label{tab:nice-single}
\resizebox{\textwidth}{!}{
\begin{tabular}{lccccc}
\toprule
Model & Threshold & Micro-F1 & Macro-F1 & Hamming Loss & LRAP \\
\midrule
Label-frequency baseline & 0.05 & 0.2221 & 0.2166 & 0.8751 & 0.4183 \\
TF-IDF Logistic Regression & 0.45 & 0.6299 & 0.5266 & 0.0901 & 0.7861 \\
TF-IDF Linear SVM & 0.45 & 0.6212 & 0.5205 & 0.0877 & 0.7843 \\
Sentence-BERT + LR & 0.50 & \textbf{0.6667} & \textbf{0.6049} & 0.0901 & \textbf{0.8286} \\
Quantum-inspired Projection & 0.85 & 0.6076 & 0.5618 & 0.0980 & 0.7694 \\
Contrastive Quantum Projection & 0.50 & 0.4907 & 0.3885 & \textbf{0.0870} & 0.6818 \\
Hybrid Quantum--SVM Fusion & 0.45 & 0.6284 & 0.5286 & \textbf{0.0870} & 0.7779 \\
\bottomrule
\end{tabular}
}
\end{table}

On this single split, Sentence-BERT obtains the highest Macro-F1 and LRAP. The
positive projection variant performs better than the rectified contrastive
variant, while the hybrid quantum--SVM model and the contrastive quantum
projection share the lowest Hamming loss on this split.
Because the split contains only 381 usable instances, these single-split
results should be interpreted cautiously.

\subsection{5-Fold Cross-validation on NICE}
Table \ref{tab:nice-cv} reports 5-fold cross-validation results with mean and
standard deviation across folds, including the fine-tuned DistilBERT baseline.

\begin{table}[htbp]
\centering
\caption{Mean and standard deviation of 5-fold cross-validation results on the NICE multi-label dataset. Fine-tuned DistilBERT is reported with its validation-selected per-label thresholds; the other rows use the global threshold protocol. Sentence-BERT achieves the highest LRAP; the hybrid quantum--SVM model obtains the highest Micro-F1 and Macro-F1 among non-embedding models, while TF-IDF Logistic Regression obtains the lowest Hamming loss.}
\label{tab:nice-cv}
\resizebox{\textwidth}{!}{
\begin{tabular}{lcccc}
\toprule
Model & Micro-F1 & Macro-F1 & Hamming Loss & LRAP \\
\midrule
Sentence-BERT + LR & 0.6493 $\pm$ 0.0357 & \textbf{0.6121 $\pm$ 0.0450} & 0.0919 $\pm$ 0.0200 & \textbf{0.8361 $\pm$ 0.0225} \\
Hybrid Quantum--SVM Fusion & \textbf{0.6839 $\pm$ 0.0331} & 0.6084 $\pm$ 0.0294 & 0.0768 $\pm$ 0.0074 & 0.8040 $\pm$ 0.0266 \\
TF-IDF Linear SVM & 0.6821 $\pm$ 0.0405 & 0.6049 $\pm$ 0.0336 & 0.0764 $\pm$ 0.0094 & 0.8056 $\pm$ 0.0262 \\
TF-IDF Logistic Regression & 0.6787 $\pm$ 0.0463 & 0.5977 $\pm$ 0.0229 & \textbf{0.0759 $\pm$ 0.0140} & 0.7990 $\pm$ 0.0240 \\
Fine-tuned DistilBERT & 0.5420 $\pm$ 0.0569 & 0.5333 $\pm$ 0.0591 & 0.1402 $\pm$ 0.0361 & 0.7105 $\pm$ 0.0516 \\
Contrastive Quantum Projection & 0.4300 $\pm$ 0.0783 & 0.4118 $\pm$ 0.0275 & 0.2074 $\pm$ 0.1634 & 0.6968 $\pm$ 0.0503 \\
Quantum-inspired Projection & 0.5931 $\pm$ 0.0166 & 0.5490 $\pm$ 0.0237 & 0.1096 $\pm$ 0.0213 & 0.7830 $\pm$ 0.0304 \\
Label-frequency baseline & 0.2328 $\pm$ 0.0034 & 0.2083 $\pm$ 0.0032 & 0.7927 $\pm$ 0.0019 & 0.4059 $\pm$ 0.0110 \\
\bottomrule
\end{tabular}
}
\end{table}

Under standard 5-fold cross-validation with a global threshold, Sentence-BERT
is the strongest Macro-F1 and LRAP baseline. The hybrid quantum--SVM fusion model is the
strongest non-embedding model in Micro-F1 and Macro-F1 and improves over both
pure quantum-inspired variants. A Wilcoxon signed-rank test over fold-level
Macro-F1 finds no significant difference between the hybrid model and TF-IDF
Linear SVM ($p=0.1875$) or TF-IDF Logistic Regression ($p=0.4375$). Given the
small number of folds and the dependence induced by cross-validation
resampling, these tests should be interpreted as descriptive evidence rather
than a strong statistical conclusion \cite{demsar2006statistical}.
The fine-tuned DistilBERT row is weaker than the frozen Sentence-BERT+LR
baseline and the TF-IDF/hybrid models in this small, imbalanced setting,
indicating that direct fine-tuning of a general-purpose encoder is not
automatically superior to strong linear baselines under the present data
budget.

\subsection{5-Fold Cross-validation with Per-label Thresholds}
Table \ref{tab:nice-per-label} reports the 5-fold results when each label
receives its own threshold selected on the validation split.

\begin{table}[htbp]
\centering
\caption{Mean and standard deviation of 5-fold results with per-label threshold calibration on the NICE multi-label dataset. Sentence-BERT has the highest Macro-F1 and LRAP, while the hybrid quantum--SVM model remains the strongest non-embedding model in Macro-F1.}
\label{tab:nice-per-label}
\resizebox{\textwidth}{!}{
\begin{tabular}{lcccc}
\toprule
Model & Micro-F1 & Macro-F1 & Hamming Loss & LRAP \\
\midrule
Sentence-BERT + LR & \textbf{0.6386 $\pm$ 0.0280} & \textbf{0.5990 $\pm$ 0.0276} & \textbf{0.0947 $\pm$ 0.0092} & \textbf{0.8361 $\pm$ 0.0225} \\
Hybrid Quantum--SVM Fusion & 0.6178 $\pm$ 0.0749 & 0.5854 $\pm$ 0.0341 & 0.1027 $\pm$ 0.0476 & 0.8040 $\pm$ 0.0266 \\
TF-IDF Linear SVM & 0.6085 $\pm$ 0.0588 & 0.5640 $\pm$ 0.0309 & 0.1057 $\pm$ 0.0378 & 0.8056 $\pm$ 0.0262 \\
TF-IDF Logistic Regression & 0.6219 $\pm$ 0.0588 & 0.5629 $\pm$ 0.0532 & 0.0967 $\pm$ 0.0195 & 0.7990 $\pm$ 0.0240 \\
Fine-tuned DistilBERT & 0.5420 $\pm$ 0.0569 & 0.5333 $\pm$ 0.0591 & 0.1402 $\pm$ 0.0361 & 0.7105 $\pm$ 0.0516 \\
\bottomrule
\end{tabular}
}
\end{table}

With per-label threshold calibration, Sentence-BERT again obtains the best
Macro-F1, and the hybrid quantum--SVM model remains the strongest
non-embedding model in Macro-F1. Absolute Macro-F1 decreases for every model
relative to the global-threshold protocol (Table~\ref{tab:nice-cv}), likely
reflecting the smaller per-label validation subsets used to select each
threshold; the hybrid model's relative drop (-3.8\%) is smaller than TF-IDF
Linear SVM's (-6.8\%) or TF-IDF Logistic Regression's (-5.8\%), suggesting the
hybrid model is comparatively more robust to per-label threshold calibration
rather than benefiting from it in absolute terms.

\subsection{Ablation Study}
Table \ref{tab:nice-ablation} reports a 5-fold ablation study of the main
quantum-inspired components. The ablation compares the original positive
projection, contrastive projection with and without interference, an SVM-only
component using the same sublinear TF-IDF representation as the hybrid model,
and hybrid fusion with different quantum weights.

\begin{table}[htbp]
\centering
\caption{Ablation study on the NICE multi-label dataset. Hybrid fusion with
$\alpha \in \{0.15, 0.30, 0.50\}$ performs within one standard deviation across
all four metrics, indicating that the projection score is useful as a
calibrated auxiliary signal but that Macro-F1 is not sensitive to the exact
quantum weight in this range; $\alpha=0.30$ is reported as the main
configuration.}
\label{tab:nice-ablation}
\resizebox{\textwidth}{!}{
\begin{tabular}{lcccc}
\toprule
Variant & Micro-F1 & Macro-F1 & Hamming Loss & LRAP \\
\midrule
Hybrid $\alpha=0.15$ & \textbf{0.6860 $\pm$ 0.0312} & \textbf{0.6091 $\pm$ 0.0293} & \textbf{0.0756 $\pm$ 0.0072} & \textbf{0.8085 $\pm$ 0.0280} \\
Hybrid $\alpha=0.30$ & 0.6839 $\pm$ 0.0331 & 0.6084 $\pm$ 0.0294 & 0.0768 $\pm$ 0.0074 & 0.8040 $\pm$ 0.0266 \\
Hybrid $\alpha=0.50$ & 0.6814 $\pm$ 0.0404 & 0.6054 $\pm$ 0.0299 & 0.0795 $\pm$ 0.0113 & 0.8012 $\pm$ 0.0232 \\
SVM-only sublinear TF-IDF & 0.6713 $\pm$ 0.0553 & 0.5945 $\pm$ 0.0409 & 0.0843 $\pm$ 0.0263 & 0.8081 $\pm$ 0.0291 \\
Positive projection without interference & 0.5948 $\pm$ 0.0238 & 0.5526 $\pm$ 0.0285 & 0.1108 $\pm$ 0.0233 & 0.7839 $\pm$ 0.0309 \\
Positive projection with interference & 0.5931 $\pm$ 0.0166 & 0.5490 $\pm$ 0.0237 & 0.1096 $\pm$ 0.0213 & 0.7830 $\pm$ 0.0304 \\
Contrastive projection without interference & 0.5447 $\pm$ 0.0420 & 0.5397 $\pm$ 0.0398 & 0.1339 $\pm$ 0.0100 & 0.7183 $\pm$ 0.0523 \\
Contrastive projection with interference & 0.4300 $\pm$ 0.0783 & 0.4118 $\pm$ 0.0275 & 0.2074 $\pm$ 0.1634 & 0.6968 $\pm$ 0.0503 \\
\bottomrule
\end{tabular}
}
\end{table}

The ablation supports four observations. First, interference provides no
measurable benefit for either projection variant on this dataset: it leaves
the positive projection essentially unchanged (Macro-F1 0.5526 without
interference versus 0.5490 with it) and substantially worsens the contrastive
projection (0.5397 without versus 0.4118 with). The best hybrid configuration
also uses a small interference weight ($\lambda=0.03$), and the fusion weight
that matters is $\alpha$, not $\lambda$. We therefore treat interference as a
secondary, largely inactive component in the current TF-IDF setting rather
than as a primary contribution, and report it mainly because the underlying
label-coupling formulation in Proposition~\ref{prop:interference} is
independently interesting. Second, the rectified contrastive
projection is semantically safer than squaring signed negative evidence, but in
this dataset the pure contrastive variant performs worse than the positive
projection. Third, the hybrid model improves over pure projection variants,
showing the value of discriminative calibration. Compared with the SVM-only
sublinear TF-IDF component, Hybrid $\alpha=0.30$ improves mean Macro-F1 from
0.5945 to 0.6084, an absolute gain of 0.0139 attributable to adding the
projection score as an auxiliary signal. Fourth, performance is best when the
projection score is used as an auxiliary signal rather than as the dominant
score.

\subsection{Explainability Evaluation}
We evaluate explanation faithfulness using the ERASER framework metrics: Comprehensiveness and Sufficiency. Across 158 true label assignments in the test split, for the contrastive projection model:
(i) Comprehensiveness (score drop when deleting top-3 terms) was 0.0305, which is 5.21 times higher than random deletion (0.0059).
(ii) Sufficiency is computed as $f(x)_c-f(x_{\mathrm{top}\text{-}3})_c$, so lower values are better because the rationale-only input should preserve the original score. The contrastive projection obtains -0.0395, whereas keeping 3 random terms drops the score by 0.0337 (ratio -1.17). This negative value means the top-3 rationale alone receives a slightly higher score than the full requirement; we interpret it cautiously as evidence that non-rationale tokens can add background noise, not as a human-rationale quality guarantee.
For the TF-IDF Linear SVM baseline:
(i) Comprehensiveness was 0.0867 (4.01 times random drop of 0.0216).
(ii) Sufficiency was 0.0484 (0.42 times random drop of 0.1141).
Relative to a random-deletion control, the intrinsic quantum-inspired explanations are meaningfully faithful and of a similar order to the SVM coefficient baseline; in absolute score-drop terms, the SVM explanation removes more score mass (0.0867 vs.\ 0.0305), so this comparison should be read as competitive on a relative, random-normalized basis rather than as uniformly stronger.

Table \ref{tab:qualitative} gives a qualitative example from the same deletion
test. The highest-contribution terms for the Security label correspond to access
control concepts that are also plausible human rationales.

\begin{table}[htbp]
\centering
\caption{Example requirement with top token-level contributions for the true NFR label, illustrating the intrinsic explanation produced by the projection model.}
\label{tab:qualitative}
\begin{tabular}{p{0.52\textwidth}p{0.16\textwidth}p{0.22\textwidth}}
\toprule
Requirement excerpt & Label & Top terms \\
\midrule
``The product shall ensure that it can only be accessed by authorized users'' & Security & authorized, only, access \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Auxiliary PROMISE-expanded Result}
The auxiliary single-label PROMISE-expanded experiment suggests that the
quantum-inspired projection model can be competitive in a single-label
setting: on one 70/30 train/test split, it obtains 0.7025 accuracy and 0.6748
Macro-F1, compared with 0.6899 accuracy and 0.6344 Macro-F1 for TF-IDF
Logistic Regression. This result comes from a single split without
cross-validation or a significance test, so, like the single-split NICE
result above, it should be interpreted cautiously; it is not the main
evidence for the multi-label research claim, and because PROMISE-expanded is
single-label, it does not exercise the multi-label interference mechanism.

\section{Discussion}
The experiments show a useful separation between accuracy and explanation.
On the main NICE multi-label dataset, Sentence-BERT is the strongest baseline
in Macro-F1 and LRAP, while the hybrid quantum--SVM model is competitive with
TF-IDF baselines and provides intrinsic token-level contributions. The pure
quantum-inspired projections are not competitive classifiers by themselves:
they lose to TF-IDF Linear SVM in all 5 cross-validation folds, with large
paired effect sizes (Cohen's $d_z=-1.36$ for positive projection and $-5.92$
for contrastive projection; Wilcoxon $p=0.0625$, the smallest value attainable
at $n=5$ folds).

Hybrid fusion is the component that makes the approach practically usable.
Its Macro-F1 is directionally higher than an SVM-only baseline sharing the
same TF-IDF representation ($+0.0139$, paired $d_z=0.49$), but the gain is
not statistically reliable: it is negative in 2 of 5 folds, and all main
Wilcoxon comparisons in Tables~\ref{tab:nice-cv} and
\ref{tab:nice-per-label} are non-significant ($p\geq0.1875$). Detecting an
effect of this size with conventional power would require roughly 30--35
paired folds, which is not realistic for one 381-instance dataset; hence the
accuracy gain should be read as suggestive, not confirmed.

To reduce reliance on fold-level significance testing, we also ran a paired
bootstrap over the held-out NICE test requirements. The hybrid model's Macro-F1
advantage over TF-IDF Linear SVM was 0.0081 with a 95\% bootstrap interval of
[-0.0100, 0.0263], while its advantage over TF-IDF Logistic Regression was
0.0020 with interval [-0.0205, 0.0227]. Both intervals overlap zero,
corroborating the cross-validation result: at the current dataset size, the
data cannot distinguish the hybrid model's accuracy from that of TF-IDF
baselines. The one accuracy comparison that is both large and well-supported
is not between the hybrid model and classical baselines, but between the
hybrid model and the pure quantum-inspired projection it fuses ($+0.140$
Macro-F1, 95\% CI $[0.074, 0.213]$, $p<0.001$): the quantum projection is too
weak to deploy on its own and depends on the SVM component to be usable,
rather than the reverse.

Although Sentence-BERT obtains the highest Macro-F1 and LRAP, the hybrid
projection model occupies a different point in the tradeoff between accuracy
and interpretability. It does not add a confirmed accuracy advantage over
classical baselines. Sentence-BERT relies on a pretrained transformer encoder
and produces dense embeddings whose individual dimensions are not directly
interpretable; explanations therefore require post-hoc methods such as LIME
\cite{ribeiro2016lime} or SHAP \cite{lundberg2017shap}. In contrast, the
projection model yields token-level contributions intrinsic to the scoring
function and passes a deletion-based faithfulness check, while remaining
statistically indistinguishable from TF-IDF baselines. The method is therefore
most attractive when intrinsic interpretability and a small dependency
footprint matter more than maximum Macro-F1, not as an accuracy improvement
in its own right.

\section{Threats to Validity}
\textbf{Dataset size and availability.} NICE contains 381 usable multi-label
NFR requirements after filtering, so all cross-validation and ablation results
come from one small corpus. The observed ranking may depend on NICE's
vocabulary, label distribution, and annotation conventions, and a second
multi-label NFR corpus is needed before making broad accuracy-generalization
claims. Multi-label NFR datasets with fine-grained subclasses are scarce; as a
partial mitigation, we include a single-label PROMISE-expanded sanity check,
but it does not exercise the multi-label interference mechanism. The current
evidence is sufficient for the paper's interpretability and diagnostic claims,
but not for a state-of-the-art accuracy claim beyond NICE.

\textbf{Baseline scope.} The current implementation includes TF-IDF baselines,
a frozen Sentence-BERT encoder, and a cross-validated fine-tuned DistilBERT
baseline. It does not include NoRBERT, RoBERTa, ML-kNN, or the
small-language-model classifier from the NICE paper. The DistilBERT result
reduces the risk that the comparison relies only on frozen embeddings, but it
is not an exhaustive neural baseline study: different pretrained encoders,
longer schedules, or domain-specific requirements models could change the
ranking. This does not affect the paper's central claims, which are explicitly
framed around interpretability and diagnostic ablation rather than
state-of-the-art accuracy.

\textbf{Model simplicity.} The quantum-inspired models are based on TF-IDF
semantic states, projection directions, and score fusion. They do not yet use
contextual neural embeddings inside the projection component.

\textbf{Threshold sensitivity.} Multi-label classification depends strongly on
decision thresholds. The experiments show that per-label threshold calibration
can change the relative ranking of models.

\textbf{Explainability evaluation.} The deletion test measures faithfulness to
the model's own scoring function, but it does not measure human usefulness or
agreement with expert rationales.

\textbf{Ablation scope.} The ablation study examines projection type,
interference, and fusion weight, but it does not yet test contextual embeddings
inside the quantum-inspired projection component.

\section{Conclusion}
This paper presented an intrinsically explainable quantum-inspired formulation
for multi-label NFR classification. Requirements are semantic states, labels
are projection directions, and predictions are explained through semantic
amplitude contributions rather than post-hoc approximations. The theoretical
analysis clarifies exactly when the model behaves like cosine-style centroid
ranking and when interference or row-wise normalization creates cross-label
coupling.

Empirically, the method's strongest role is explanatory and diagnostic. The
hybrid quantum--SVM model is competitive with TF-IDF baselines and is the
strongest non-embedding model in Macro-F1, but its accuracy advantage is not
statistically significant at the current dataset size; pure projection variants
are weaker as standalone classifiers. The deletion-based evaluation shows that
top contribution terms are substantially more influential than random terms,
supporting the faithfulness of the intrinsic explanations. Thus, the paper
contributes a transparent multi-label scoring framework and a careful account
of when its quantum-inspired geometry helps, not an inflated accuracy claim.

Future work should integrate contextual embeddings into the quantum-inspired
representation and compare with additional fine-tuned BERT/RoBERTa
\cite{liu2019roberta} and NoRBERT baselines under broader tuning budgets. It
should also evaluate ML-kNN \cite{zhang2007mlknn} as a classical multi-label
baseline, test the approach on a second multi-label NFR corpus, and collect
human rationale judgments for explanation quality.

\section*{Reproducibility}
The implementation contains scripts for reproducing the experiments:
\begin{itemize}
    \item \texttt{scripts/run\_nice\_multilabel\_experiment.py}
    \item \texttt{scripts/run\_nice\_cv\_experiment.py}
    \item \texttt{scripts/run\_nice\_per\_label\_threshold\_experiment.py}
    \item \texttt{scripts/run\_nice\_ablation\_experiment.py}
    \item \texttt{scripts/run\_explainability\_deletion\_test.py}
    \item \texttt{scripts/run\_nice\_robustness\_experiment.py}
    \item \texttt{scripts/run\_nice\_finetuned\_transformer\_experiment.py}
    \item \texttt{scripts/run\_all\_experiments.py}
    \item \texttt{scripts/run\_promise\_experiment.py}
\end{itemize}
For anonymous review, the artifact is available at the anonymized repository:
\url{https://anonymous.4open.science/r/nfr-review-artifact-CE2D/}.

\section*{Use of AI Tools}
During the preparation of this manuscript, the authors used a large language
model as an assistant for implementation checks, dependency diagnosis,
experiment re-runs, statistical-script review, consistency checks, and
language editing. The propositions, statistical interpretations, contribution
framing, and final manuscript text were independently reviewed and rewritten
by the authors against the code, logs, and exported result files. The authors
take full responsibility for the experiments, proofs, analyses, and claims in
the paper.

\bibliographystyle{splncs04}
\bibliography{references}

\end{document}
